Ab Initio Effective One-Electron Potential Operators: Applications for Charge-Transfer Energy in Effective Fragment Potentials

### 2019-12-08 papers/2_oep_technique.CT/paper-r3.tex
Revised paper uploaded

\input{header.tex}
\begin{document}
\preprint{AIP/123-OEP}

@@

\begin{abstract}
The concept of the effective one\hyp{}electron potentials (OEP) has been useful for many decades
in efficient description of electronic structure of chemical systems, especially extended
molecular aggregates such as interacting molecules in condensed phases.
Here, the general method for effective elimination of the electron repulsion integrals (ERI)
based on OEP's is presented, that is tuned towards the fragment\hyp{}based calculation methodologies
such as the second generation of the effective fragment potentials (EFP2) method. 
Two general types of the OEP operator matrix elements
are distinguished and treated either via the distributed multipole expansion
or the extended density fitting schemes developed in this work. 
The OEP technique was then applied
to address the problem of using usually incomplete EFP2 settings in most of
applications in interaction energy and 
molecular dynamics simulations 
due to prohibitively high computational cost of evaluating the
charge transfer (CT) effects as compared to other effects.
The alternative OEP\hyp{}based CT energy model is proposed
in the context of the
intermolecular perturbation theory with Hartree\hyp{}Fock uninteracting gas\hyp{}phase 
reference wavefunctions, compatible with the EFP2 formulation.
It is found that the computational cost can be reduced
up to 20 times as compared to the CT energy method within the EFP2 scheme
without compromising the accuracy for a wide range of weakly interacting 
neutral molecular complexes. 
Therefore, it is believed that
the proposed model can be used in calculations within 
the EFP2 framework, making the CT energy term no longer the bottleneck
in EFP2\hyp{}based simulations of complex systems.
\end{abstract}

@@

Charge transfer (CT) between molecules occurs when the net electronic populations 
of interacting molecules change which leads to an additional stabilization 
of a molecular aggregate.\cite{Otto.Ladik.IJQC.1980,Stone.TheTheoryOfIntermolecularForces.1996}
Although CT processes can be relatively easily scrutinized based
on the total amount of electronic charge transferred between interacting species,\cite{Otto.Ladik.IJQC.1980}
evaluation of its contribution to the intermolecular interaction
potential\cite{Stone.TheTheoryOfIntermolecularForces.1996} is far from trivial due to 
its notably complex quantum mechanical (QM) origins 
even at the Hartree\hyp{}Fock\cite{Roothaan.RevModPhys.1951} 
approximation level,\cite{Otto.Ladik.ChemPhys.1975}
and cannot be realized in terms of any classical nor semi\hyp{}classical approach.
In fact, CT does not naturally emerge in the symmetry 
adapted perturbation theory\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} (SAPT)
due to technical aspects related with the use of finite basis set expansions
in modern quantum chemistry calculations and avoiding the basis set superimposition error (BSSE). 
Stone and Misquitta showed that CT energy
can be extracted from SAPT calculations by comparing the induction energies
of fictional systems in which basis functions are centered either on the monomers only, or the entire 
interacting complex.\cite{Stone.Misquitta.CPL.2009}
CT energy was formulated by Murrell et al.~\cite{Murrell.Randic.Williams.Longuet-Higgins.ProcRSocLondA.1965} 
in their perturbation theory in the region of small wavefunction overlap up to second order.
However, all these theories are computationally too expensive to apply for efficient
calculation of intermolecular forces in molecular dynamics because
they involces instantaneous calculations of the electron repulsion integrals (ERI's)
and their four\hyp{}index transformation to molecular orbital (MO) basis.\cite{Otto.Ladik.ChemPhys.1975}

The apparent difficulty in theoretically characterising the CT energy
in terms of the interacting molecular fragments
is indeed a challenge in developing modern force fields 
or \emph{ab initio} fragmentation methods\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}
for modeling structure
and dynamics in condensed phases.
This needs to be contrasted with the Coulombic electrostatics,
non\hyp{}Coulombic repulsion (i.e., due to Pauli exclusion principle), 
dispersion and induction for instance, 
which to a certain extent
can be well described by, respectively, 
the multipole moments of charge density distributions,\cite{Sokalski.Poirier.CPL.1983,
Etchebest.Lavery.Pullman.TheorChimActa.1982,Stone.JCTC.2005}
the van der Waals repulsive and attractive potentials, 
as well as the polarizability models, i.e., all being relatively simple and computationally
inexpensive to evaluate mathematical models. 
Due to this reason, the CT effects 
%while usually of significant and non\hyp{}negligible magnitude as compared 
%to the total interaction energy, 
are not explicitly included in any molecular mechanics force field developed up to date.
%such as OPLS, Amber, CHARMM, Gromos or even polarizale AMOEBA force field. 
The only way to
incorporate the CT effects in the condensed\hyp{}phase simulations,
apart from performing full QM electronic structure simulations,
is through the second generation of the effective fragment potential 
(EFP2) method,\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013,
   Nguyen.Pachter.Day.JCP.2014,
   Day.Jensen.Gordon.Webb.Stevens.Krauss.Garmer.Basch.Cohen.JCP.1996}
developed by Gordon group.\cite{Li.Gordon.Jensen.JCP.2006,Xu.Gordon.JCP.2013}
EFP2 is an \emph{ab initio} force field
derived on the grounds of the first\hyp{}principles perturbation theory
at the HF level and including the intermolecular dispersion effects
by the response theory. 
%based on the imaginary frequency\hyp{}dependent dipole\hyp{}dipole polarizabilities.
That is to say, the total intermolecular 
interaction potential, which can also be perceived as an approximation
to SAPT at HF level with added intermolecular dispersion correction, is approximated as 
%
\begin{equation}\label{e:u-efp2}
 U^{\rm EFP2} \approx U^{\rm Coul} + U^{\rm Ex-Rep} + U^{\rm Ind} + U^{\rm Disp} + U^{\rm CT} \;,
\end{equation}
%
where $U^{\rm Coul}$ is the Coulombic interaction energy of the unperturbed charge\hyp{}density
distributions of the monomers, approximated by the distributed multipole approximation,
$U^{\rm Ex-Rep}$ is the exchange\hyp{}repulsion energy originating from the Pauli exclusion
principle, that could be roughly understood as the Heitler\hyp{}London interaction
energy with $U^{\rm Coul}$ subtracted, $U^{\rm Ind}$ and $U^{\rm Disp}$
are the induction and dispersion energies obtained from the distributed polarizability
approximation, and finally $U^{\rm CT}$ is the CT energy, the focus of this work.

Despite the considerable success of the EFP2 theory
in accurately modeling the extended molecular systems 
like solutions and recently even biomolecules
with the level of accuracy
roughly arguably comparable to the M{\o}ller\hyp{}Plesset perturbation theory,
evaluation of the CT energy in EFP2 model is still too costly for
typical uses in the
molecular dynamics simulations. It has been reported that 
the implementation of the EFP2 CT energy component in GAMESS
is on average 3-10 times more demanding than the exchange\hyp{}repulsion
component, becoming a bottleneck of evaluation of total interaction 
energy and gradients.\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013,
Li.Gordon.Jensen.JCP.2006}
In effect, the CT energy component is usually ignored in most of the applications. 
As a noteworthy example, 
CT term were not implemented in the official release 
of the recent LIBEFP library for linking quantum chemistry packages 
with the EFP2 functionalities.\cite{Kaliman.Slipchenko.JCC.2015}

@@

The effect of CT on the interaction energy 
is known to be often non\hyp{}negligible, especially
in donor\hyp{}acceptor systems such as H\hyp{}bonded species
and charged complexes, although highly depends on the theoretical
approach being used, i.e., the choice of the reference
wavefunctions.
The approach adopted by the founders of EFP2 model
is based on gas\hyp{}phase reference (unperturbed) HF wavefunctions,
%The most common methodology,
%adopted also by the founders of EFP2 model,
%is 
%which is based on the isolated monomeric wave functons
%as references 
here broadly referred to as the CT/HF0 level of theory.
In this formulation, the CT energy can be understood
as a part of the polarization energy that cannot be
linked to dispersion and pure induction interactions. 
In technical words, mixing of virtual and occupied MO's,
that are assigned according to certain criteria 
to different unperturbed wavefunctions, occurs
unlike the pure induction and dispersion, in which such mixings are contained
solely within one monomer.
%This level of definition of the CT energy will be referred here as to
%the IPTO-CT energy. 
%For defining the CT/HF0, 
%most common reference is the HF approximation
%most certainly because the HF energy functional is known
%and its properties are well understood.
%In the EFP2 model, the total interaction energy is approximated as
%a sum of five contributions due to the Coulombic interaction of the unperturbed (isolated)
%charge densities of the monomers, exchange\hyp{}repulsion energy,
%induction energy due to mutual polarization, intermolecular dispersion energy,
%as well as the CT energy.
%For developing the efficient 
%\emph{ab initio} fragmentation method 
%
% the second generation of the effective fragment potential method (EFP2),
%the Hartree\hyp{}Fock (HF) model or the density functional theory (DFT) were
%employed as starting points 
%Modelling the structure and dynamics of extended molecular aggregates, such as
%liquids and solutions require
%EFP2 relies on the intermolecular 
%perturbation theory, and the total interaction energy is approximated as ...
%
%\begin{equation}
% U^{\rm EFP2} \approx U^{\rm Coul} + U^{\rm Ex-Rep} + U^{\rm Ind} + U^{\rm Disp} + U^{\rm CT} \;.
%\end{equation}
%%

One of the main goals of this work, apart from developing
more efficient model of the CT/HF0 energy that
is compatible with the EFP2 energetics, is to
extend the effective one\hyp{}electron potentials (here referred to as the OEP) 
technique
widely
used previously\cite{Roothaan.RevModPhys.1951,
Hohenberg.Kohn.PhysRev.1964,
Kohn.Sham.PhysRev.1965,
Otto.Ladik.ChemPhys.1975,
Holas.March.PhysRevA.1991,
Weber.Thiel.TCA.2000,
Neese.JCP.2005,
Cisneros.Andres.Piquemal.Darden.JCP.2005,
Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Li.Gordon.Jensen.JCP.2006,
Blasiak.Lee.Cho.JCP.2013,
Blasiak.Maj.Cho.Gora.JCTC.2015}
to simplify the rigorous and costly quantum chemistry models of extended systems
with a particular emphasis on solvation phenomena and molecular dynamics.
The presented OEP technique of removing ERI's from the working
equations
follows the notion of the importance of 
one\hyp{}electron densities in chemistry\cite{Kohn.Sham.PhysRev.1965,Holas.March.PhysRevA.1991},
thus reducing the complicated summations involving ERI's to much shorter expressions involving
only one\hyp{}electron integrals (OEI's).
Therefore, the OEP computational method is first presented in Section~\ref{s:2.oep}.
Next, in Section~\ref{s:3.ct},
the new technique is used to derive an alternative formulation of the CT/HF0 energy
compatible with the EFP2 method. Subsequently, after details of computations
are discussed in Section~\ref{s:4.calculations}, in Section~\ref{s:5.results} the validation of the 
OEP\hyp{}based CT/HF0 model is presented and its performance in terms 
of accuracy and computational speed is compared against the EFP2
model. Finally, few concluding remarks and outlook of future work
are given in Section~\ref{s:6.conclusions}.
%ERI's, which are oftentimes the main bottleneck
%in simulating large systems, will be projected out from the working equations
%to ultimately form the effective fragments in a form of small matrices that are much easier to handle.
%The proposed methodology aims in removing ERI's from the working equations
%to ultimately form the effective fragments that are much easier to handle.
%
%It is believed that the proposed concept can be helpful in designing future methods tuned 
%for studying electronically excited states in solutions, better describing molecular
%dynamics as well as enabling the vibrational analysis of extended systems
%In this work, we propose an alternative formulation of the CT energy compatible with the
%EFP2 method. We start from the perturbation theory of Murrell et al.
%and
%introduce the effective one\hyp{}electron potential (OEP) operators
%to eliminate the electron\hyp{}repulsion integrals (ERI's) from the original theory.
%Later we validate the model against the EFP2 method in Section XXX. Finally,
%we conclude this work and provide few remarks on future development in Section XXX.

@@

This scheme is most applicable in the case of matrix elements of the type
$
 \tBraKet{j^B}{\hat{v}_{\rm eff}^A}{l^B}
$
because it can be considered as a Coulombic interaction between $\rho^B_{jl}$
and $\rho^A_{\rm eff}$. These matrix elements require ERI's of type
$\BraKet{AA}{BB}$ only.
In general, given certain two effective one\hyp{}electron density distributions,
the associated effective Coulombic interaction energy can be estimated from the classical formula
according to
%
\begin{equation} \label{e:coulomb-energy:general}
 \Delta^{(XY)}_{\rm eff} E_{\rm Coul} = \iint \frac{\rho_{\rm eff}^X({\bf r}_1) \rho_{\rm eff}^Y({\bf r}_2)}
 {\vert {\bf r}_1 - {\bf r}_2 \vert} 
d{\bf r}_1 d{\bf r}_2 \;.
\end{equation}
%
The above integral can be approximated by 
applying the DMTP expansion 
to both effective potential operators
and dropping the `$\hat{}$' symbol in the multipole operators
which leads to\cite{Stone.TheTheoryOfIntermolecularForces.1996}
%
\begin{multline} \label{e:coulomb-energy:dmtp}
  \Delta^{(XY)}_{\rm eff} E_{\rm Coul} \approx
 \rho_{\rm eff}^X \odot \rho_{\rm eff}^Y \equiv
 \sum_{u\in A} \sum_{w\in B} \Big\{ 
 T^{(0)}_{uw}
 q_{\rm eff}^{(u)}  q_{\rm eff}^{(w)} \\
 - {\bf T}^{(1)}_{uw} \cdot 
   \left[ q_{\rm eff}^{(u)} {\BM\upmu}_{\rm eff}^{(w)} - q_{\rm eff}^{(w)} {\BM\upmu}_{\rm eff}^{(u)} \right]
 - {\bf T}^{(2)}_{uw} : 
  {\BM\upmu}_{\rm eff}^{(u)}  \otimes {\BM\upmu}_{\rm eff}^{(w)} 
 \ldots
 \Big\}
\end{multline}
%
The symbol `$\odot$' denotes the sum of all the tensor contractions
performed over the DMTP's of molecule $X$ and $Y$ to yield the associated interaction energy.
The choice of the distribution centres as well as the truncation order of the multipole expansion
is crucial in compromising the accuracy and computational cost of the resulting expressions.
There is many ways in which this can be achieved, e.g., through the distributed multipole analysis (DMA)
of Stone\cite{Stone.JCTC.2005},
the cumulative atomic multipole moments (CAMM) of Sokalski and Poirier\cite{Sokalski.Poirier.CPL.1983},
the localised distributed multipole expansion (LMTP) of Etchtebest, Lavery and Pullman\cite{Etchebest.Lavery.Pullman.TheorChimActa.1982}
or schemes based on fitting to electrostatic potential such as ChelpG\cite{Breneman.Wiberg.JCC.1990} for instance.

>>>>

\documentclass[aip,jcp,amsmath,amssymb,reprint,floatfix]{revtex4-1}
\input{header.tex}
\begin{document}
\preprint{AIP/123-OEP}

@@

\begin{abstract}
The concept of the effective one\hyp{}electron potentials (OEP) has been useful for many decades
in efficient description of electronic structure of chemical systems, especially extended
molecular aggregates such as interacting molecules in condensed phases.
Here, a general method for effective OEP\hyp{}based elimination of electron repulsion integrals (ERI),
that is tuned
towards the fragment\hyp{}based calculation methodologies
such as the second generation of the effective fragment potentials (EFP2) method,
is presented.
Two general types of the OEP operator matrix elements
are distinguished and treated either via the distributed multipole expansion
or the extended density fitting schemes developed in this work. 
The OEP technique is then applied
to address the problem of using incomplete EFP2 settings in many
applications in interaction energy and 
molecular dynamics simulations 
due to prohibitively high computational cost of evaluating the
charge transfer (CT) effects as compared to other effects.
The alternative OEP\hyp{}based CT energy model is proposed
in the context of the
intermolecular perturbation theory with Hartree\hyp{}Fock uninteracting gas\hyp{}phase 
reference wavefunctions, compatible with the EFP2 formulation.
It is found that the computational cost can be reduced
up to 20 times as compared to the CT energy method within the EFP2 scheme
without compromising the accuracy for a wide range of weakly interacting 
neutral molecular complexes. 
Therefore, it is believed that
the proposed model can be used within 
the EFP2 framework, making the CT energy term no longer the bottleneck
in EFP2\hyp{}based simulations of complex systems.
\end{abstract}

@@

Charge transfer (CT) between molecules occurs when the net electronic populations 
of interacting molecules change which leads to an additional stabilization 
of a molecular aggregate.\cite{Otto.Ladik.IJQC.1980,Stone.TheTheoryOfIntermolecularForces.1996}
Although CT processes can be relatively easily scrutinized based
on the total amount of electronic charge transferred between interacting species,\cite{Otto.Ladik.IJQC.1980}
evaluation of its contribution to the intermolecular interaction
potential\cite{Stone.TheTheoryOfIntermolecularForces.1996} is far from trivial due to 
its notably complex quantum mechanical (QM) origins 
even at the Hartree\hyp{}Fock\cite{Roothaan.RevModPhys.1951} (HF)
approximation level,\cite{Jensen.JCP.2001,Otto.Ladik.ChemPhys.1975}
and cannot be realized in terms of any classical nor semi\hyp{}classical approach.
In fact, CT does not naturally emerge in the symmetry 
adapted perturbation theory\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} (SAPT)
due to technical aspects related with the use of finite basis set expansions
in modern quantum chemistry calculations and avoiding the basis set superimposition error (BSSE). 
Stone and Misquitta showed that CT energy
can be extracted from SAPT calculations by comparing the induction energies
of fictional systems in which basis functions are centered either on the monomers only, or the entire 
interacting complex.\cite{Stone.Misquitta.CPL.2009}
CT energy was formulated by Murrell et al.~\cite{Murrell.Randic.Williams.Longuet-Higgins.ProcRSocLondA.1965} 
in their perturbation theory in the region of small wavefunction overlap up to second order.
However, all these theories are computationally too expensive to apply for efficient
calculation of intermolecular forces in molecular dynamics because
they involve calculation of the electron repulsion integrals (ERI's)
and their four\hyp{}index transformation to molecular orbital (MO) basis.\cite{Otto.Ladik.ChemPhys.1975}

The apparent difficulty in theoretically characterising the CT energy
in terms of the interacting molecular fragments
is indeed a challenge in developing modern force fields 
or \emph{ab initio} fragmentation methods\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}
for modeling structure
and dynamics in condensed phases.\cite{Demerdash.Mao.Liu.Head-Gordon.Head-Gordon.JCP.2017}
This needs to be contrasted with the Coulombic electrostatics,
non\hyp{}Coulombic repulsion (i.e., due to Pauli exclusion principle), 
dispersion and induction for instance, 
which to a certain extent
can be well described by, respectively, 
the distributed multipole moments of charge densities,\cite{Sokalski.Poirier.CPL.1983,
Etchebest.Lavery.Pullman.TheorChimActa.1982,Stone.JCTC.2005}
the van der Waals repulsive and attractive potentials, 
as well as the polarizability models, i.e., all being relatively simple and computationally
inexpensive to evaluate mathematical models. 
Due to this reason, the CT effects 
%while usually of significant and non\hyp{}negligible magnitude as compared 
%to the total interaction energy, 
are not explicitly included in any molecular mechanics force field 
developed up to date.\cite{Demerdash.Yap.Head-Gordon.AnnuRevPhysChem.2014}
%such as OPLS, Amber, CHARMM, Gromos or even polarizale AMOEBA force field. 
The only way to
incorporate the CT effects in the condensed\hyp{}phase simulations
in \emph{ab initio} manner,
apart from performing full QM electronic structure simulations
or non\hyp{}force\hyp{}field\hyp{}based 
fragmentation techniques\cite{Leverentz.Maerzke.Keasler.Siepmann.Truhlar.PCCP.2012,
Dahlke.Truhlar.JCTC.2008},
is through the second generation of the effective fragment potential 
(EFP2) method,\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013,
   Nguyen.Pachter.Day.JCP.2014,
   Day.Jensen.Gordon.Webb.Stevens.Krauss.Garmer.Basch.Cohen.JCP.1996,
   Sattasathuchana.Xu.Gordon.JPCA.2019,
   Blasiak.Londergan.Webb.Cho.ACR.2017,
   Kuroki.Mori.ChemLett.2016,
   Ghosh.Cho.Choi.JPCB.2014}
developed by Jensen, Gordon and co\hyp{}workers.\cite{Jensen.JCP.2001,Li.Gordon.Jensen.JCP.2006,Xu.Gordon.JCP.2013}
EFP2 is an \emph{ab initio} force field
derived on the grounds of the first\hyp{}principles %perturbation theory
at the HF level\cite{Jensen.Gordon.MolPhys.1996,Jensen.Gordon.JCP.1998,Li.Netzloff.Gordon.JCP.2006} 
and including the intermolecular dispersion effects
by the response theory.\cite{Adamovic.Gordon.MolPhys.2005,Xu.Zahariev.Gordon.JCTC.2014}
%based on the imaginary frequency\hyp{}dependent dipole\hyp{}dipole polarizabilities.
That is to say, the total intermolecular 
interaction potential, which can also be perceived as an approximation
to SAPT at HF level with added intermolecular dispersion correction, is approximated as 
%
\begin{equation}\label{e:u-efp2}
 U^{\rm EFP2} \approx U^{\rm Coul} + U^{\rm Ex-Rep} + U^{\rm Ind} + U^{\rm Disp} + U^{\rm CT} \;,
\end{equation}
%
where $U^{\rm Coul}$ is the Coulombic interaction energy of the unperturbed charge\hyp{}density
distributions of the monomers, treated by the distributed multipole approximation
with damping to account for the charge\hyp{}penetration effects,\cite{Slipchenko.Gordon.JCC.2007}
$U^{\rm Ex-Rep}$ is the exchange\hyp{}repulsion energy originating from the Pauli exclusion
principle,\cite{Jensen.Gordon.MolPhys.1996,Jensen.Gordon.JCP.1998} 
that could be roughly understood as the Heitler\hyp{}London interaction
energy\cite{Chalasinski.Gutowski.MolPhys.1985} with $U^{\rm Coul}$ subtracted, $U^{\rm Ind}$ and $U^{\rm Disp}$
are the induction and dispersion energies obtained from the distributed polarizability
approximation,\cite{Li.Netzloff.Gordon.JCP.2006,Adamovic.Gordon.MolPhys.2005,Xu.Zahariev.Gordon.JCTC.2014} 
and finally $U^{\rm CT}$ is the CT energy,\cite{Li.Gordon.Jensen.JCP.2006,Xu.Gordon.JCP.2013} 
the focus of this work.

Despite the considerable success of the EFP2 theory
in accurately modeling the extended molecular systems 
like solutions\cite{Sattasathuchana.Xu.Gordon.JPCA.2019,
   Blasiak.Londergan.Webb.Cho.ACR.2017,
   Kuroki.Mori.ChemLett.2016,
   Ghosh.Cho.Choi.JPCB.2014}
and recently even biomolecules\cite{Blasiak.Ritchie.Webb.Cho.PCCP.2016,
Xu.Blasiak.Cho.Layfield.Londergan.JPCL.2018,
Gurunathan.Acharya.Ghosh.Kosenkov.Kaliman.Shao.Krylov.Slipchenko.JPCB.2016,
Ghosh.Kosenkov.Vanovschi.Williams.Herbert.Gordon.Schmidt.Slipchenko.Krylov.JPCA.2010}
with the level of accuracy
reaching in many cases\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
the M{\o}ller\hyp{}Plesset perturbation theory\cite{Moller.Plesset.PhysRev.1934},
evaluation of the CT energy in EFP2 model is still relatively costly for
typical uses in the
molecular dynamics simulations. It has been reported that 
the implementation of the EFP2 CT energy and gradient in GAMESS US computer program\cite{GAMESS.JCC.1993}
is on average 30--40 times more demanding than the exchange\hyp{}repulsion
component, becoming a bottleneck of evaluation of total interaction 
energy and gradients.\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013,
Li.Gordon.Jensen.JCP.2006}
In effect, the CT energy component is often ignored 
in many applications.\cite{Kuroki.Mori.JPCB.2019,
Blasiak.Londergan.Webb.Cho.ACR.2017,
Tommaso.Lafiosca.Cappelli.JCTC.2017,
Gurunathan.Acharya.Ghosh.Kosenkov.Kaliman.Shao.Krylov.Slipchenko.JPCB.2016,
Budzak.Laurent.Laurence.Medved.Jacquemin.JCTC.2016,
Smith.Gordon.Slipchenko.JPCA.2011,
Smith.Gordon.Slipchenko.JPCA.2011.2,
Ghosh.Kosenkov.Vanovschi.Williams.Herbert.Gordon.Schmidt.Slipchenko.Krylov.JPCA.2010,
Smith.Gordon.Slipchenko.JPCA.2008}
In fact,
EFP2 CT term is available only in the GAMESS US quantum chemistry program,\cite{GAMESS.JCC.1993}
whereas it is neither supported in the official release 
of the recent LIBEFP library for linking quantum chemistry packages 
with the EFP2 functionalities,\cite{Kaliman.Slipchenko.JCC.2015}
nor in the Q\hyp{}CHEM quantum chemistry 
program,\cite{Ghosh.Kosenkov.Vanovschi.Flick.Kaliman.Shao.Gilbert.Krylov.Slipchenko.JCC.2013}
contrary to electrostatic, exchange\hyp{}repulsion, induction and dispersion EFP2 terms.

@@

The effect of CT on the interaction energy 
is known to be often non\hyp{}negligible, especially
in donor\hyp{}acceptor systems such as H\hyp{}bonded species
and charged complexes,\cite{Devarajan.Gaenko.Gordon.Windus.Fragmentation.2017,
Devarajan.Windus.Gordon.JPCA.2011} 
although highly depends on the theoretical
approach being used, i.e., the choice of the reference
wavefunctions.
The approach adopted by the founders of EFP2 model
is based on gas\hyp{}phase reference (unperturbed) HF wavefunctions,
%The most common methodology,
%adopted also by the founders of EFP2 model,
%is 
%which is based on the isolated monomeric wave functons
%as references 
which will be here broadly referred to as the CT/HF0 level of theory
and the symbol `0' denotes unperturbed wavefunctions.
In this formulation, the CT energy can be understood
as a part of the polarization energy that cannot be
linked to dispersion and pure induction interactions. 
In technical words, mixing of virtual and occupied MO's,
that are assigned according to certain criteria 
to different unperturbed wavefunctions, occurs
unlike the pure induction and dispersion, in which such mixings are contained
solely within one monomer.\cite{Stone.TheTheoryOfIntermolecularForces.1996}
%This level of definition of the CT energy will be referred here as to
%the IPTO-CT energy. 
%For defining the CT/HF0, 
%most common reference is the HF approximation
%most certainly because the HF energy functional is known
%and its properties are well understood.
%In the EFP2 model, the total interaction energy is approximated as
%a sum of five contributions due to the Coulombic interaction of the unperturbed (isolated)
%charge densities of the monomers, exchange\hyp{}repulsion energy,
%induction energy due to mutual polarization, intermolecular dispersion energy,
%as well as the CT energy.
%For developing the efficient 
%\emph{ab initio} fragmentation method 
%
% the second generation of the effective fragment potential method (EFP2),
%the Hartree\hyp{}Fock (HF) model or the density functional theory (DFT) were
%employed as starting points 
%Modelling the structure and dynamics of extended molecular aggregates, such as
%liquids and solutions require
%EFP2 relies on the intermolecular 
%perturbation theory, and the total interaction energy is approximated as ...
%
%\begin{equation}
% U^{\rm EFP2} \approx U^{\rm Coul} + U^{\rm Ex-Rep} + U^{\rm Ind} + U^{\rm Disp} + U^{\rm CT} \;.
%\end{equation}
%%

One of the main goals of this work, apart from developing
more efficient model of the CT/HF0 energy that
is compatible with the EFP2 energetics, is to
extend the one\hyp{}electron effective potentials (here referred to as the OEP) 
technique
widely
used previously\cite{Roothaan.RevModPhys.1951,
Hohenberg.Kohn.PhysRev.1964,
Kohn.Sham.PhysRev.1965,
Otto.Ladik.ChemPhys.1975,
Holas.March.PhysRevA.1991,
Weber.Thiel.TCA.2000,
Neese.JCP.2005,
Cisneros.Andres.Piquemal.Darden.JCP.2005,
Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Li.Gordon.Jensen.JCP.2006,
Blasiak.Lee.Cho.JCP.2013,
Blasiak.Maj.Cho.Gora.JCTC.2015}
to simplify the rigorous and costly quantum chemistry models of extended systems
with a particular emphasis on solvation phenomena and molecular dynamics.
The presented OEP technique of removing ERI's from the working
equations
follows the notion of the importance of 
one\hyp{}electron densities in chemistry\cite{Kohn.Sham.PhysRev.1965,Holas.March.PhysRevA.1991},
thus reducing the complicated summations involving ERI's to much shorter expressions involving
only one\hyp{}electron integrals (OEI's).
Therefore, the OEP computational method is first presented in Section~\ref{s:2.oep}.
Next, in Section~\ref{s:3.ct},
the new technique is used to derive an alternative formulation of the CT/HF0 energy
compatible with the EFP2 method. Subsequently, after details of computations
are discussed in Section~\ref{s:4.calculations}, in Section~\ref{s:5.results} the validation of the 
OEP\hyp{}based CT/HF0 model is presented and its performance in terms 
of accuracy and computational speed is compared against the EFP2
model. Finally, few concluding remarks and outlook of future work
are given in Section~\ref{s:6.conclusions}.
%ERI's, which are oftentimes the main bottleneck
%in simulating large systems, will be projected out from the working equations
%to ultimately form the effective fragments in a form of small matrices that are much easier to handle.
%The proposed methodology aims in removing ERI's from the working equations
%to ultimately form the effective fragments that are much easier to handle.
%
%It is believed that the proposed concept can be helpful in designing future methods tuned 
%for studying electronically excited states in solutions, better describing molecular
%dynamics as well as enabling the vibrational analysis of extended systems
%In this work, we propose an alternative formulation of the CT energy compatible with the
%EFP2 method. We start from the perturbation theory of Murrell et al.
%and
%introduce the effective one\hyp{}electron potential (OEP) operators
%to eliminate the electron\hyp{}repulsion integrals (ERI's) from the original theory.
%Later we validate the model against the EFP2 method in Section XXX. Finally,
%we conclude this work and provide few remarks on future development in Section XXX.

@@

This scheme is most applicable in the case of matrix elements of the type
$
 \tBraKet{j^B}{\hat{v}_{\rm eff}^A}{l^B}
$
because it can be considered as a Coulombic interaction between $\rho^B_{jl}$
and $\rho^A_{\rm eff}$. These matrix elements require ERI's of type
$\BraKet{AA}{BB}$ only.
In general, given certain two effective one\hyp{}electron density distributions,
the associated effective Coulombic interaction energy can be estimated from the classical formula
according to
%
\begin{equation} \label{e:coulomb-energy:general}
 \Delta^{(XY)}_{\rm eff} E_{\rm Coul} = \iint \frac{\rho_{\rm eff}^X({\bf r}_1) \rho_{\rm eff}^Y({\bf r}_2)}
 {\vert {\bf r}_1 - {\bf r}_2 \vert} 
d{\bf r}_1 d{\bf r}_2 \;.
\end{equation}
%
The above integral can be approximated by 
applying the DMTP expansion 
to both effective potential operators
and dropping the $\hat{}$ symbol in the multipole operators
which leads to\cite{Stone.TheTheoryOfIntermolecularForces.1996}
%
\begin{multline} \label{e:coulomb-energy:dmtp}
  \Delta^{(XY)}_{\rm eff} E_{\rm Coul} \approx
 \rho_{\rm eff}^X \odot \rho_{\rm eff}^Y \equiv
 \sum_{u\in A} \sum_{w\in B} \Big\{ 
 T^{(0)}_{uw}
 q_{\rm eff}^{(u)}  q_{\rm eff}^{(w)} \\
 - {\bf T}^{(1)}_{uw} \cdot 
   \left[ q_{\rm eff}^{(u)} {\BM\upmu}_{\rm eff}^{(w)} - q_{\rm eff}^{(w)} {\BM\upmu}_{\rm eff}^{(u)} \right]
 - {\bf T}^{(2)}_{uw} : 
  {\BM\upmu}_{\rm eff}^{(u)}  \otimes {\BM\upmu}_{\rm eff}^{(w)} 
 \ldots
 \Big\}
\end{multline}
%
The symbol `$\odot$' denotes the sum of all the tensor contractions
performed over the DMTP's of molecule $X$ and $Y$ to yield the associated interaction energy.
The choice of the distribution centres as well as the truncation order of the multipole expansion
is crucial in compromising the accuracy and computational cost of the resulting expressions.
There is many ways in which this can be achieved, e.g., through the distributed multipole analysis (DMA)
of Stone and Alderton\cite{Stone.Alderton.MolPhys.1985,Stone.JCTC.2005},
the cumulative atomic multipole moments (CAMM) of Sokalski and Poirier\cite{Sokalski.Poirier.CPL.1983},
the localised distributed multipole expansion (LMTP) of Etchtebest, Lavery and Pullman\cite{Etchebest.Lavery.Pullman.TheorChimActa.1982}
or schemes based on fitting to electrostatic potential such as ChelpG\cite{Breneman.Wiberg.JCC.1990} for instance.
